---
title: "Children Greenland"
author: Me
date: '`r format(Sys.Date(), "%d. %b %Y")`'
slug: Children_Greenland
categories:
  - data analysis
tags: [data analysis]
subtitle: ''
summary: ''
authors: []
lastmod: '2025-07-22T14:27:06Z'
featured: no
image:
  caption: 'created by DALL·E'
  focal_point: ''
  preview_only: no
projects: []
output: 
  bookdown::html_document2
draft: TRUE
bibliography: biblio_children.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=F, message=F}
library(readxl)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(ggthemes)
library(stringr)
library(paletteer)
```

```{r, echo=F}
path_data <- "/Users/solon/Desktop/Website/content/post/children_greenland/"

Inuit <- read_excel(paste0(path_data, "Inuit-Norse-toys_v5.xlsx"), sheet = "Inuit") %>%
   mutate(Culture = "Inuit")

Norse <- read_excel(paste0(path_data, "Inuit-Norse-toys_v5.xlsx"), sheet = "Norse") %>% 
   mutate(Culture = "Norse")
```

```{r, echo = F, warning=F}
select_cols <- c("Culture", "Society", "Item_name", "Type", "No.", "Dates", 
                 "Material_metal", "Material_plant", "Material_stone", "Material_bone", "Material_other", "Material_description",
                 "MaxLength_mm", "MaxWidth_mm", "MaxThickness_mm", "MaxHeight_mm", "Finder")

Inuit_sel <- Inuit %>% select(all_of(select_cols)) %>% na.omit()
Norse_sel <- Norse %>% select(all_of(select_cols)) %>% na.omit()

df <- rbind(Inuit_sel, Norse_sel) %>%
    mutate(No. = as.numeric(No.)) %>%
    mutate(
        # Extract numeric start of the range
        Year_start = as.numeric(sub("-.*", "", Dates)),
        # Make Dates an ordered factor by Year_start
        Dates = factor(Dates, levels = unique(Dates[order(Year_start)])),
        # Form Material count
        Material_count = rowSums(across(c(Material_metal, Material_plant, Material_stone, Material_bone, Material_other), 
                                        function(x) x == "YES")),
        across(
            c(MaxLength_mm, MaxWidth_mm, MaxThickness_mm, MaxHeight_mm),
            ~ as.numeric(ifelse(. == "-" | . == "", NA, .)))
        )
```

When we think about survival in harsh climates, images of hunters, warriors, and explorers often come to mind. But in Greenland during the Little Ice Age (14th century AD), survival wasn’t just about adults—it involved children too. Recent research reveals that the toys and play objects of Inuit and Norse children weren’t just for fun; they were tools of adaptation, creativity, and resilience [@meyer2025playing]. By examining the variety, materials, and complexity of these objects, archaeologists are uncovering how play shaped the skills and innovative thinking necessary to thrive in one of the world’s most challenging environments. In this post, I'm exploring how something as simple as a child’s toy can tell profound stories about culture and survival. 

Data from @meyer2025playing. 

# Toys as Training: Comparing Norse and Inuit Childhoods in Greenland

The Inuit entered Greenland from Canada from AD 1000. The Norse, on the other hand, reached Greenland from Iceland around the same time.

The Norse and the Inuit developed very different ways of interacting with their environment, shaped by their unique social structures and ecological relationships. These differences are reflected even in the toys children played with.

During the period when the Norse and Inuit coexisted in Greenland (roughly AD 1000–1400), their approaches to childhood play diverged in meaningful ways (Figure \@ref(fig:fig1)). Inuit society emphasized toys that prepared children for life in the Arctic. Miniature hunting tools and social play objects were common, likely serving as practice for essential survival skills. In contrast, Norse children’s toys were more closely tied to agriculture, reflecting the early introduction of farm work in their daily lives.

Among the Inuit, little change in these categories is observable over time. For the Norse, the variation is more pronounced, although this pattern may simply reflect the limited sample size. These differences highlight how childhood play can offer a window into broader cultural strategies for survival and adaptation.

```{r fig1, echo=F, fig.cap = "Proportions of toy categories over time."}
df_types <- df %>% 
    mutate(No. = as.numeric(No.)) %>% 
    group_by(Culture, Dates, Type) %>%
    # Sum the number of items ("No.") per group
    summarise(total_items = sum(No., na.rm = TRUE), .groups = "drop_last") %>%
    # Calculate total per Culture + Dates
    mutate(total_culture_date = sum(total_items),
           proportion = total_items / total_culture_date) %>%
    # Sort neatly
    arrange(Culture, Dates, desc(proportion))

df_types %>% filter(Dates == "1000-1200" | Dates == "1200-1400") %>% 
    ggplot(., aes(x = Dates, y = proportion*100, col = Type, group = Type)) +
    geom_point() +
    geom_line() +
    geom_text_repel(data = df_types %>% filter(Dates == "1000-1200"), 
                    aes(label = Type)) +
    labs(x = "Time", y = "Proportion, %") +
    facet_wrap(.~ Culture) +
    scale_color_brewer(palette = "Dark2") +
    theme_hc(12) +
    theme(legend.position = "none")
```
Looking at the individual items, dolls and figures (figurines) made up the largest percentage of the Inuit and Norse toys, respectively (Figure \@ref(fig:fig2)).

It is interesting to notice the diversity of items within each category for the Inuits compared to the Norse. This diversity is commonly seen as an indication of innovation [@kuhn2020evolution]. 

```{r fig2, echo=F, warning=F, fig.cap = "Proportions of toy items over time."}
df_items <- df %>% 
    group_by(Culture, Dates, Item_name) %>%
    # Sum the number of items ("No.") per group
    summarise(Type = first(Type), 
              total_items = sum(No., na.rm = TRUE), .groups = "drop_last") %>%
    # Calculate total per Culture + Dates
    mutate(total_culture_date = sum(total_items),
           proportion = total_items / total_culture_date) %>%
    # Sort neatly
    arrange(Culture, Dates, desc(proportion))

df_items %>% filter(Dates == "1000-1200" | Dates == "1200-1400")  %>% 
    ggplot(., aes(x = Dates, y = proportion*100, col = Type, group = Item_name)) +
    geom_point() +
    geom_line() +
    labs(x = "Time", y = "Proportion, %") + 
    geom_text_repel(data = df_items %>% filter(Dates == "1000-1200"), 
                    aes(label = Item_name)) +
    facet_wrap(.~ Culture) +
    #theme_minimal() +
    scale_color_brewer(palette = "Dark2") +
    theme_hc(12) 
```
Another indication of diversity comes from the materials used. Usually only one material is used per item with only 6 items using combinations of two materials (bone and metal or wood and metal), all made by Inuit. 

For both cultures, the most prevalent material was wood (Figure \@ref(fig:fig3)). Followed by bone for the Inuit and stone for the Norse. The bones mainly originated from sea animals such as seals and whales [e.g. @seersholm2016dna]. 

So what?

```{r fig3, echo=F, fig.cap = "Proportions of toy materials over time."}
material_cols <- c("Material_metal", "Material_plant", "Material_stone", "Material_bone",
                   "Material_other")

material_prop <- df %>%
    # Convert "YES"/"NO" to 1/0 for counting
    mutate(across(all_of(material_cols), ~ ifelse(. == "YES", 1, 0))) %>%
    # Group by Culture and Dates
    group_by(Culture, Dates) %>%
    # Sum number of YES for each material and total items
    summarise(
        #total_items = n(), #n(), #n(),  # total rows (or use sum(No.) if you want actual number of items)
        across(all_of(material_cols), sum, .names = "sum_{col}"),
        .groups = "drop") %>%
    # Calculate proportion for each material
   mutate(total_items_calc = rowSums(across(starts_with("sum_")))) %>%
   mutate(across(starts_with("sum_"), ~ . / total_items_calc, .names = "prop_{col}")) %>%
    select(Culture, Dates, starts_with("prop_")) %>%
    tidyr::pivot_longer(
        cols = starts_with("prop_sum_"),     # columns with proportions
        names_to = "Material",               # new column for material type
        values_to = "Proportion"             # new column for proportion value
    ) %>%
    # Optional: clean up the Material names
    mutate(Material = str_to_sentence(gsub("prop_sum_Material_", "", Material)),
           Material = ifelse(Material == "Plant", "Wood", Material))
    
material_prop %>% 
   filter(Dates == "1000-1200" | Dates == "1200-1400") %>% 
    ggplot(., aes(x = Dates, y = Proportion*100, group = Material, col = Material)) +
    facet_wrap(Culture ~ .) +
    geom_point() +
    geom_line() +
    geom_text_repel(data = material_prop %>% filter(Dates == "1000-1200"), 
                    aes(label = Material)) +
    labs(x = "Time", y = "Proportion, %") + 
    theme_hc(12) +
    scale_color_paletteer_d("PrettyCols::Bold") +
    theme(legend.position = "none")
```
```{r, echo=F, eval = F}
df %>% filter(Material_bone == "YES") %>% filter(Culture == "Inuit") %>%
   filter(Dates == "1000-1200" | Dates == "1200-1400") %>% 
   reframe(unique(Material_description)) 
```

# Shifting Play: Inuit Toys and Cultural Change Over 800 Years

For the Inuit, the dataset spans a period of approximately 800 years, allowing us to examine changes over an extended timeframe.

The proportion of social play items (such as dolls) declines up to the 1600–1800 period, after which it increases sharply (Figures \@ref(fig:fig4) and \@ref(fig:fig5)). In contrast, the proportion of weapon-related items rises until the 1600–1800s and then begins to decline. Tool-related artifacts show a steadier, more gradual increase across the sequence.

The most pronounced shifts appear to occur between 1600 and 1800—a period corresponding to renewed contact between the Inuit and Europeans.

```{r fig4, echo=F, warning = F, message=F, fig.cap = "Proportion of toy categories for Inuit."}
df_types %>% 
   filter(Culture == "Inuit") %>% 
   ggplot(., aes(x = Dates, y = proportion*100, col = Type, group = Type)) +
   geom_point() +
   geom_line() +
   #geom_text_repel(aes(label = Type)) +
   labs(x = "Time", y = "Proportion, %") + 
   geom_text_repel(data = df_types %>% 
                      filter(Culture == "Inuit") %>% 
                      filter(Dates == "1000-1200"), 
                    aes(label = Type)) +
   facet_wrap(.~ Culture) +
   scale_color_brewer(palette = "Dark2") +
   theme_hc(12) +
   theme(legend.position = "none")

# 10 most prevalent items in a period
item_label <- df_items %>% filter(Culture == "Inuit") %>% 
   #filter(Dates == "1400-1600") %>%
   filter(Dates == ">1800") %>%
   arrange(desc(proportion)) %>%
   top_n(10) %>%
   mutate(Dates = "1000-1200")
```

```{r fig5, fig.cap = "Proportion of toy items for Inuit.", echo=F, warning=F}
df_items %>% 
   filter(Culture == "Inuit") %>% 
    ggplot(., aes(x = Dates, y = proportion*100, col = Type, group = Item_name)) +
    geom_point() +
    geom_line() +
    #geom_text_repel(aes(label = Item_name)) +
    facet_wrap(.~ Culture) +
    labs(x = "Time", y = "Proportion, %") + 
    geom_text_repel(data = item_label, 
                    aes(label = Item_name)) +
    facet_wrap(.~ Culture) +
    #theme_minimal() +
    scale_color_brewer(palette = "Dark2") +
    theme_hc(12) +
    theme(legend.position = "none")
```

Among the recorded toy dimensions, the average toy length shows the most pronounced changes over time (Figure \@ref(fig:fig6)). This pattern may reflect shifts in design and function, variations in material availability, or a combination of both. Notably, between AD 1400 and 1600 there is a marked decline in the proportion of wooden toys, accompanied by an increase in bone objects (Figure \@ref(fig:fig7)). However, when material use is examined by toy type, this reduction in wooden items appears to be driven primarily by tools and weapons (Figure \@ref(fig:fig8)).

```{r fig6, echo=F, warning=F, message = F, fig.cap = "Toy dimensions for Inuit."}
avg_dimensions <- df %>%
    group_by(Culture, Dates, Type) %>%
    summarise(
        mean_length = mean(MaxLength_mm, na.rm = TRUE),
        mean_width = mean(MaxWidth_mm, na.rm = TRUE),
        mean_thickness = mean(MaxThickness_mm, na.rm = TRUE),
        mean_height = mean(MaxHeight_mm, na.rm = TRUE),
        n_items = n()
    ) %>%
    ungroup()

avg_dimensions_long <- avg_dimensions %>%
    tidyr::pivot_longer(
        cols = starts_with("mean_"),     # the dimension columns
        names_to = "Dimension",          # new column for the variable names
        values_to = "Value"            # new column for the values
    ) %>%
    mutate(
        Dimension = gsub("mean_", "", Dimension),  # clean up names (optional)
        Dimension = tools::toTitleCase(Dimension)  # make them look nice: "Length", "Width", etc.
    )
avg_dimensions_long %>% 
   filter(Culture == "Inuit") %>% 
   ggplot(., aes(x = Dates, y = Value, col = Type, group = Type)) +
   geom_point() +
   geom_line() +
   labs(y = "Average Millimeters", x = "Time", col = "") +
   #geom_text_repel(aes(label = Type)) +
   facet_grid(. ~ Dimension, scales = "free") +
   scale_color_brewer(palette = "Dark2") +
   theme_hc(12) +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```
```{r fig7, fig.cap = "Proportion of material for Inuit.", echo=F, warning=F}
material_prop %>% 
   filter(Culture == "Inuit") %>% 
   #filter(Dates == "1000-1200" | Dates == "1200-1400") %>% 
    ggplot(., aes(x = Dates, y = Proportion*100, group = Material, col = Material)) +
    #facet_wrap(Culture ~ .) +
    geom_point() +
    geom_line() +
    geom_text_repel(data = material_prop %>% 
                       filter(Dates == "1000-1200") %>%
                       distinct(Dates, Material, .keep_all = T), 
                    aes(label = Material), max.overlaps = 1) +
    labs(x = "Time", y = "Proportion, %") + 
    theme_hc(12) +
    scale_color_paletteer_d("PrettyCols::Bold") +
    theme(legend.position = "none")
```

Interestingly, each toy category exhibits distinct material preferences (Figure \@ref(fig:fig8)). Skill-based play objects were often made of bone, social play items (such as dolls) and transport toys (boats, kayaks) were predominantly crafted from wood, while tools were mainly produced from stone—especially after AD 1200. These patterns suggest that variations in toy dimensions and materials are more likely linked to changes in design or function rather than to material scarcity.

Weapons present an interesting case: their frequency increases in AD 1600-1800 (Figure \@ref(fig:fig4)), they become more diverse (Figure \@ref(fig:fig9)), and the dominant material shifts from bone to wood (Figure \@ref(fig:fig8)). This trend has been hypothesized to reflect broader changes in hunting implements that emerged in response to the increasingly harsh environmental conditions of the time (cite paper).

```{r fig8, echo=F, warning = F, message = F, fig.cap = "Proportion of material per toy category for Inuit."}
material_prop <- df %>%
    # Convert "YES"/"NO" to 1/0 for counting
    mutate(across(all_of(material_cols), ~ ifelse(. == "YES", 1, 0))) %>%
    # Group by Culture and Dates
    group_by(Culture, Dates, Type) %>%
    # Sum number of YES for each material and total items
    summarise(
        #total_items = sum(No., na.rm = T),  # total rows (or use sum(No.) if you want actual number of items)
        across(all_of(material_cols), sum, .names = "sum_{col}"),
        .groups = "drop") %>%
    # Calculate proportion for each material
   mutate(total_items_calc = rowSums(across(starts_with("sum_")))) %>%
    mutate(across(starts_with("sum_"), ~ . / total_items_calc, .names = "prop_{col}")) %>%
    select(Culture, Dates, Type, starts_with("prop_")) %>%
    tidyr::pivot_longer(
        cols = starts_with("prop_sum_"),     # columns with proportions
        names_to = "Material",               # new column for material type
        values_to = "Proportion"             # new column for proportion value
    ) %>%
    # Optional: clean up the Material names
    mutate(Material = str_to_sentence(gsub("prop_sum_Material_", "", Material)),
           Material = ifelse(Material == "Plant", "Wood", Material))

material_prop %>% 
   filter(Culture == "Inuit") %>% 
   ggplot(., aes(x = Dates, y = Proportion*100, group = Material, col = Material)) +
   geom_point() +
   geom_line() +
   geom_text_repel(data = material_prop %>% 
                      filter(Dates == "1000-1200") %>%
                      distinct(Dates, Material, .keep_all = T), 
                    aes(label = Material)) +
   facet_wrap(Type ~ ., scales = "free") +
   labs(x = "Time", y = "Proportion, %") + 
   theme_hc(12) +
   scale_color_paletteer_d("PrettyCols::Bold") +
   theme(legend.position = "none") +
   theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r, fig9, echo=F, fig.cap="Number of distinct weapon toys over time."}
df_distinct_weapons <- df_items %>% 
   filter(Culture == "Inuit") %>% 
   filter(Type == "Weapon") %>%
   group_by(Dates, Type, Culture) %>%
   reframe(N_distinct = length(unique(Item_name)))
   
df_distinct_weapons %>% 
   ggplot(., aes(x = Dates, y = N_distinct, 
                 fill = Type, group = Type)) +
   geom_bar(stat = "identity") +
    #geom_text_repel(aes(label = Item_name)) +
   labs(x = "Time", y = "Number of Distinct Weapon Toys") + 
   facet_wrap(.~ Culture) +
   scale_fill_brewer(palette = "Dark2") +
   theme_hc(12) +
   theme(legend.position = "none")
```


## References 


