---
title: "Missed Opportunity? Rethinking Quality of Life Analysis in the SONIA Trial"
author: Me
date: '`r format(Sys.Date(), "%d. %b %Y")`'
slug: sonia_trial
categories:
  - reporting
tags: [reporting]
subtitle: ''
summary: 'To add summary'
authors: []
lastmod: '2025-12-20T14:27:06Z'
featured: no
image:
  caption: 'created by DALL·E'
  focal_point: ''
  preview_only: no
projects: []
draft: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=F}
library(ggplot2)
```

The [SONIA trial](https://www.nature.com/articles/s41586-024-08035-2) made headlines by challenging a major assumption in breast cancer care: that powerful drugs like CDK4/6 inhibitors must be used right away in advanced HR+/HER2– breast cancer. The study found **no significant difference in survival** between using these drugs early or later. It even showed that later use led to fewer side effects and lower costs.

But there’s one area where the results are … less satisfying.

The authors reported **no difference in health-related quality of life (HRQOL)** between the early and delayed treatment groups, based on the widely used FACT-B questionnaire. At first glance, this seems reassuring—patients felt just as good either way.

But let’s look a little closer.

## The HRQOL Analysis Was Likely Too Simple

The authors used a main effects model to compare the FACT-B scores over time. This approach makes two key assumptions:

1. The only difference between the groups is a constant shift—i.e., one group is consistently better or worse than the other across time.
    - In their model, the coefficient for the "CDK4/6i-second" group was −0.91, implying a minor, consistent shift.
2. FACT-B scores change linearly over time.

Both assumptions are strong—and likely inadequate.

```{r}
intercept <- 37.950
arm_second_group <- -0.91
time_coef <- -0.008
FACTB_baseline <- 0.676

time <- seq(0, 640, by = 10)
first_group <- intercept + (time_coef * time) + FACTB_baseline
second_group <- first_group + arm_second_group

original_df <- data.frame(y = c(first_group, second_group),
                          group = c(rep("CDK4/6 first", length(time)), rep("CDK4/6 second", length(time))),
                          time = time)
```

```{r}
ggplot(original_df) +
   geom_line(aes(y = y, x = time, colour = group))
```

## A Better Way to Ask the Question

Instead of asking, “Were average quality of life scores different across time?”, why not ask a more patient-centered question:

### **Hypothesis 1**

*Patients who start CDK4/6 inhibitors earlier (first-line group) experience a **faster decline** in quality of life compared to those who start them later.*

Why might this be true? Because the first-line group had more severe side effects—more than 1,000 additional grade 3+ events than the second-line group. It’s logical to expect this would show up in how patients feel over time.

Testing this hypothesis means including an interaction between time and treatment group. This would allow the slopes of the two groups’ HRQOL trajectories to differ.

Visually, this would look like one group’s curve dropping faster than the other.

```{r}
interaction_coeff <- -0.005
arm_second_group <- -0.9
main <- intercept + (time_coef * time) + FACTB_baseline
first_group_int <- main + (interaction_coeff*time)
second_group_int <- main + arm_second_group


hyp1_df <- data.frame(y = c(first_group_int, second_group_int),
                          group = c(rep("CDK4/6 first", length(time)), 
                                    rep("CDK4/6 second", length(time))),
                          time = time)

ggplot(hyp1_df) +
   geom_line(aes(y = y, x = time, colour = group))

```

### **Hypothesis 2**

*Even if the average scores at the end are the same, the **shapes of the trajectories** might differ—non-linear patterns, dips, or recoveries.*

Here, we’re relaxing the idea that quality of life changes in a neat, straight line. Real life (and real cancer treatment) is rarely linear. Patients often experience early/late toxicity, and plateaus. A model assuming straight-line change might completely miss that.

Pictorially, this might show two wavy lines crossing or diverging—even if they end at the same place.

```{r}

```

## Why This Matters

These more flexible, hypothesis-driven approaches might **better capture what patients actually experience**.

The SONIA trial is important, but its quality of life analysis could have gone further. With the right modeling, we might learn even more about **how and when treatment side effects truly impact patients’ lives**.

