---
title: Approximating Binomial with Poisson
author: ''
date: '2019-02-05'
slug: approximating-binomial-with-poisson
categories: [probability, R]
tags: [probability, R]
output: bookdown::html_document2
header:
  caption: ''
  image: ''
  preview: yes
link-citations: yes
bibliography: approx_bin.bib
---

It is usually taught in statistics classes that Binomial probabilities can be approximated by Poisson probabilities, which are generally easier to calculate. This approximation is valid "when $n$ is large and $np$ is small," and rules of thumb are sometimes given.

In this post I’ll walk through a simple proof showing that the Poisson distribution is really just the Binomial with $n$ (the number of trials) approaching infinity and $p$ (the probability of success in each trail) approaching zero. I'll then provide some numerical examples to investigate how good is the approximation. 

## Proof 
The Binomial distribution describes the probability that there will be $x$ successes in a sample
of size $n$, chosen with replacement from a population where the probability of success is $p$. 

Let $X \sim Binomial(n, p)$, that is 

\begin{equation}
(\#eq:binom)
   P(X = x) = {n\choose x} p^x (1-p)^{n-x},
\end{equation}

where $x= 0, 1, \dots, n$. Define the number

$$\lambda = np$$

This is the rate of success. That’s the number of trials $n$—however many there are—times the chance of success $p$ for each of those trials. If we repeat the experiment every day, we will be getting $\lambda$ successes per day on average. 

Solving for $p$, we get:

$$ p = \frac{\lambda}{n}$$
We then substitute this into \@ref(eq:binom), and take the limit as $n$ goes to infinity

$$ \lim_{n \to \infty}P(X = x) =  \lim_{n \to \infty} \frac{n!}{x!(n-x)!} \bigg( \frac{\lambda}{n} \bigg)^x \bigg( 1-\frac{\lambda}{n} \bigg)^{n-x}$$

I then collect the constants (terms that don't depend on $n$) in front and split the last term into two 

\begin{equation}
   (\#eq:limit)
   \frac{\lambda^x}{x!}  \lim_{n \to \infty} \color{blue}{\frac{n!}{(n-x)!} \bigg( \frac{1}{n} \bigg)^x} \color{red}{ \bigg( 1-\frac{\lambda}{n} \bigg)^n } \color{green}{\bigg( 1-\frac{\lambda}{n} \bigg)^{-x}}
\end{equation}

Now let’s take the limit of this right-hand side one term at a time. 

1. We start with the blue term 

$$\color{blue}{ \lim_{n \to \infty} \frac{n!}{(n-x)!} \bigg( \frac{1}{n} \bigg)^x }$$
The numerator and denominator can be expanded as follows 

$$\color{blue}{ \lim_{n \to \infty} \frac{(n)(n-1)(n-2)\dots(n-x)(n-x-1)\dots (1)}{(n-x)(n-x-1)(n-x-2)\dots (1)}\bigg( \frac{1}{n} \bigg)^x }$$

The $(n-x)(n-x-1)\dots(1)$ terms cancel from both the numerator and denominator, leaving the following

$$\color{blue}{ \lim_{n \to \infty} \frac{(n)(n-1)(n-2)(n-x+1)}{n^x} }$$
This can be rewrited as

$$\color{blue}{  \lim_{n \to \infty} \frac{n}{n} \frac{(n-1)}{n} \frac{(n-2)}{n} \frac{(n-x+1)}{n} }$$
This is because there were $x$ terms in both the numerator and denominator. Clearly, every one of these $x$ terms approaches 1 as $n$ approaches infinity. So we know this just simplifies to one. So we’re done with the first step. 

2. Now we focus on the red term of \@ref(eq:limit)

$$\color{red}{ \lim_{n \to \infty} \bigg( 1-\frac{\lambda}{n} \bigg)^n }$$

Recall the [definition](https://en.wikipedia.org/wiki/E_(mathematical_constant)) of $e= 2.7182\dots$ is 

$$ \lim_{a \to \infty} \bigg(1 + \frac{1}{a}\bigg)^a$$
Our goal here is to find a way to manipulate our expression to look more like the definition of $e$, which we know the limit of. Let’s define a number $a$ as

$$ a = -\frac{n}{\lambda}$$

Substituting it into our expression we get 

$$ \color{red}{ \lim_{n \to \infty} \bigg( 1-\frac{\lambda}{n} \bigg)^n = \lim_{n \to \infty} \bigg( 1+\frac{1}{a} \bigg)^{-a\lambda} = e^{-\lambda} }$$
So we've finished with the middle term. 

3. The third term of \@ref(eq:limit) is 

$$\color{green}{ \lim_{n \to \infty}  \bigg( 1-\frac{\lambda}{n} \bigg)^{-x} }$$
As $n$ approaches infinity, this term becomes $1^{-x}$ which is equal to one. And that takes care of our last term. 

Putting these together we can re-write \@ref(eq:limit) as 

$$ \frac{\lambda^x}{x!}  \lim_{n \to \infty} \color{blue}{ \frac{n!}{(n-x)!} \bigg( \frac{1}{n} \bigg)^x} \color{red}{ \bigg( 1-\frac{\lambda}{n} \bigg)^n} \color{green}{ \bigg( 1-\frac{\lambda}{n} \bigg)^{-x} } = \frac{\lambda^x}{x!} \color{red}{ e^{-\lambda} }$$
which is the pmf of a Poisson random variable $Y$, i.e 

$$P(Y = y)  = \frac{\lambda^y}{y!} e^{-\lambda}$$

where $y = 0, 1, \dots$. So we have shown that the Poisson distribution is a special case of the Binomial, in which the number of trials grows to infinity and the chance of success in any trial approaches zero. And that completes the proof.

@casella2002statistical provide a much shorter proof based on moment generating functions. 

A natural question is how good is this approximation? It turns out it is quite good even for moderate $p$ and $n$ as we'll see with a few numerical examples. 

## Code 
A [rule of thumb](https://www.itl.nist.gov/div898/handbook/pmc/section3/pmc331.htm) says for the approximation to be good: 

   > "The sample size $n$ should be equal to or larger than 20 and the probability of a single success, $p$, should be smaller than or equal to 0.05. If $n$ > 100, the approximation is excellent if $np$ is also < 10." 


Let's try a few scenarios. I have slightly modified the code from [here](https://www.math.utah.edu/~treiberg/M3074PoisApproxEg.pdf).

```{r}
# plots the pmfs of Binomial and Poisson
pl <- function(n, p, a, b) {
   
   clr <- rainbow(15)[ceiling(c(10.68978, 14.24863))]
   lambda <- n * p
   mx <- max(dbinom(a:b, n, p))
      
   plot(
      c(a:b, a:b),
      c(dbinom(a:b, n, p), dpois(a:b, lambda)),
      type = "n",
      main = paste("Poisson Approx. to Binomial, n=", n, ", p=", p, ", lambda=", lambda),
      ylab = "Probability",
      xlab = "x")
   points((a:b) - .15,
          dbinom(a:b, n, p),
          type = "h",
          col = clr[1],
          lwd = 10)
   points((a:b) + .15,
          dpois(a:b, lambda),
          type = "h",
          col = clr[2],
          lwd = 10)
   legend(b - 3.5, mx, legend = c("Binomial(x,n,p)", "Poisson(x,lambda)"), fill = clr, bg = "white")
}

```

I start with the recommendation: $n$ = 20, $p$ = 0.05. This gives $\lambda= 1$. Already the approximation seems reasonable.  

```{r}
pl(20, 0.05, 0, 10)
```

For $n$ = 10, $p$ = 0.3 it doesn't seem to work very well.   

```{r}
pl(10, 0.3, 0, 10)
```

But if we increase $n$ and decrease $p$ in order to come home with the same $\lambda$ value things improve. 

```{r}
pl(100, 0.03, 0, 10)
```
Lastly, for 1000 trials the distributions are indistinguishable. 

```{r}
pl(1000, 0.003, 0, 10)
```

## References