---
title: "# 224 Puzzle from the New Scientist"
author: Me
date: '2023-07-10'
slug: puzzle224
categories:
  - puzzle
tags: [puzzle]
subtitle: ''
summary: ''
authors: []
lastmod: '2023-02-08T14:27:06Z'
featured: no
image:
  caption: 'created by DALL·E'
  focal_point: ''
  preview_only: no
projects: []
draft: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[Puzzle #224](https://www.newscientist.com/article/mg25834411-000-puzzle-224-can-you-find-the-ring-inside-the-russian-dolls/) from the New Scientist is:

> I collect Russian dolls, the type where each doll can be opened to reveal a smaller one inside. I am particularly fond of my simple, single-coloured ones, which come in sets of five (and, unusually, have a hollow smallest doll). I have five lovely sets of them, each a different colour.

> Alas, while I was out, my daughter Kira rearranged them so that each large doll now contains one each of the four other colours. She proudly tells me that no blue doll contains a doll that has a yellow doll anywhere within it. There is no doll that contains a pink doll with a red doll anywhere within it. And no yellow doll contains a green doll with a pink doll anywhere within it.

> “By the way, have you seen my wedding ring?” I ask her.

> “Ah, I put that inside the smallest blue doll” replies Kira.

> Which coloured doll should l open first if I want to find the ring as quickly as possible?

```{r, fig.align="center", out.height='460px', out.width='800px', echo=FALSE}
knitr::include_graphics("./Tree_fig.pdf")
```

The figure shows all the possible arrangements of the five dolls. The letter indicates the colour: R for red, Y for yellow and so on. The index indicates the doll size: 5 for the smallest and 1 for the largest. 

As an example, the top path - `P1 R2 G3 Y4 B5`- corresponds to the combination pink, red, green, yellow and blue (from the largest to the smallest doll). 

Further, `B5` occupies the left hand size since we know the smallest doll is blue. 

With the a few lines of code I recreate the figure. Each row of the output corresponds to a set of five dolls. For example, the last row corresponds to the combination yellow, red, pink, green and blue. In total, we have 24 possible arrangements - same as in the figure.

```{r, message = F}
# pkgs used in this post
library(gtools)
library(stringi)
library(stringr)
library(rebus) 
library(purrr)

colour <- c("blue", "yellow", "pink", "red", "green") # all the different colours

numb <- 1:5 # the size index (1 is the largest)

opt <- paste0(rep(colour, each = 5), "_", numb) # combine colour and index

perms <- permutations(n = 25, r = 5, v = opt, repeats.allowed = F) # all the combinations of 5 dolls

# define function to use 
my_fun <- function(x, data, pattern = NULL){
   # x: the row index of `data`
   # data: the input dataset
   # pattern: the pattern to search for in each row of `data`
   
   if(is.null(pattern)){
      perms3 <- stri_split_fixed(data[x,], "_", simplify = T) # split the vectors into two
      dupl <- sum(stri_duplicated(perms3[,1])) # check for uniqueness
      sorted <- is.unsorted(perms3[,2], strictly = T) # check for correct order
   
      if(dupl == 0 & sorted == 0){ # if no duplicated and correct order data is returned
         out <- data[x,]
         } else {
            out <- c()
            }
      } else {
         splitted <- stri_split_fixed(data[x,], "_", simplify = T) # split the vectors into two
         ind <- str_detect(paste(splitted[,1], collapse = ""), pattern = pattern, negate = T) # find patterns
      
         if(ind == 1){ # if pattern is found it's returned
            out <- data[x,]
            }else{
               out <- c()
            }
      }
   
   return(out)
}

##### add restrictions
# 1: select only the ones where blue is the smallest doll
blue_smallest_doll <- which(perms[,5] == "blue_5") # blue is the smallest doll

perms2 <- perms[blue_smallest_doll,]

# 2: select only the combinations where each colour appears once and indexes are sorted correctly 
combinations <- matrix( # simplify final result into a matrix 
   unlist( # simplify output of map()
      map(1:nrow(perms2), my_fun, data = perms2)), # apply my_fun to all rows of perms2
   ncol = 5, byrow = TRUE) 

combinations 
```

Then, we need to take into account the two additional restrictions:

1. "no doll that contains a pink doll with a red doll anywhere within it", and
2. "no yellow doll contains a green doll with a pink doll anywhere within it".

The first one is easy. We need to discard all solutions where pink is followed by red. This gives us 12 remaining solutions. 
```{r}
# 3: apply first restriction
# specify pattern: "no doll contains a pink with a red doll anywhere within it"
pattern_pink_red <- ("pink" %R% zero_or_more(WRD) %R% "red") 

combinations_update <- matrix(
   unlist(
      map(1:nrow(combinations), my_fun, data = combinations, pattern = pattern_pink_red)), 
   ncol = 5, byrow = TRUE)

combinations_update
```

Now, we move on to the second restriction. Unfortunately, the 2nd restriction is a quite vague, leading to different interpretations and different final solutions. Here are a few examples.

### 1st interpretation 
A way to interpret the restriction is to remove all solutions where a pink doll is inside a green one, which is *at the same time*, inside a yellow one. This results in 9 solutions. 

```{r}
##### 1st interpretation 
# specify pattern: "no yellow doll contains a green doll with a pink inside it"
pattern_yellow_green_pink <- ("yellow" %R% zero_or_more(WRD) %R% "green" %R% zero_or_more(WRD) %R% "pink") 

combinations_update1 <- matrix(
   unlist(
      map(1:nrow(combinations_update), my_fun, data = combinations_update, 
          pattern = pattern_yellow_green_pink)), 
   ncol = 5, byrow = TRUE)

combinations_update1
```

### 2nd interpretation 
Another interpretation is to *additionally* remove solutions where pink is inside green, *independently* of where yellow is. This results in 4 solutions. 

```{r}
##### 2nd interpretation 
# specify pattern: "no green doll with a pink doll anywhere within it"
pattern_green_pink <- ("green" %R% zero_or_more(WRD) %R% "pink") 
                              
combinations_update2 <- matrix(
   unlist(
      map(1:nrow(combinations_update), my_fun, data = combinations_update, 
          pattern = or(pattern_yellow_green_pink, pattern_green_pink))), 
   ncol = 5, byrow = TRUE)

combinations_update2
```

### 3rd interpretation
Lastly, we can *further* remove solutions where green is inside yellow *independently* of where pink is. 

```{r}
#### 3rd interpretation
pattern_yellow_green<- ("yellow" %R% zero_or_more(WRD) %R% "green")

combinations_update3 <- matrix(
   unlist(
      map(1:nrow(combinations_update), my_fun, data = combinations_update, 
          pattern = or(pattern_yellow_green_pink, pattern_green_pink, pattern_yellow_green))), 
   ncol = 5, byrow = TRUE)

combinations_update3
```

This gives a unique solution - `R1 P2 G3 Y4 B5` - which does not seem to violate any of the restrictions. Unfortunately, it is different from the solution given by the New Scientist, which is `Y1 R2 P3 G4 B5`!!! 

