---
title: "# 131 Puzzle from the New Scientist"
author: ''
date: '2021-09-21'
slug: puzzle_131
categories: [puzzle, probability]
tags: [puzzle, probability]
output: bookdown::html_document2
header:
  caption: ''
  image: ''
  preview: yes
link-citations: yes
draft: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Solution to [#131 "The Paradise Club"](https://www.newscientist.com/article/mg25133522-800-puzzle-131-what-is-the-probability-of-winning-this-game-of-chance/#ixzz775sxHixz) puzzle from the [New Scientist](https://www.newscientist.com/). This week (No 3352 - 18 September 2021) the New Scientist posted the following puzzle: 

>Down at The Paradise Club, Gus and Bart take it in turns to roll a pair of dice. The first person to score his favourite score for two dice wins, which means being treated to a drink by the other (the loser). They each favour a different prime number as a score with two dice and it so happens that their chances of getting their favourite score is the same for each.

> What is that probability? If Gus goes first what are his chances of being bought a drink?

The maximum score (i.e. sum) when rolling two dice is 12. The prime numbers from 2 (the minimum score) to 12 are 

$$2, 3, 5, 7, 11$$

Each number can be the result of the following dice combinations 

2: {1, 1}

3: {2, 1}

5: {2, 3}, {4, 1}

7: {4, 3}, {5, 2}, {6, 1}

11: {6, 5}

For example, a score of 2, can only be achieved if both dice are {1, 1} (1 combination in total). A score of 3 can be achieved either with {2, 1} or {1, 2} (2 combinations in total).

The table below shows the total number of combinations for each score

```{r, echo=F}
Score <- c(2, 3, 5, 7, 11)
Comb <- c(1, 2, 4, 6, 2)

x <- cbind(Score, Comb)

knitr::kable(x, "simple", col.names = c("Score", "Total Combinations"), align = "cc")
```


The puzzle states that the chances of getting their favourite score is the same for  Gus and Bart. Since each combination is equiprobable[^1] the only scores that result in the same chances are 3 and 11. 

[^1]: I have assumed that each die is fair, that is the occurrence of each number is equally likely. 


Now we need to find that probability. We can calculate it as 

Probability = Number of favourable combinations/ total number of combinations

We already know the number of favourable combinations (i.e. the number of combinations that will give the win) - it is 2. There are 2 ways to get a score of 3 or 11 (see table). The total number of combinations is 36. There are 36 possible outcomes when we roll two dice (6 from the first and 6 from the second die).

Hence, the probability of getting their favourite score is $2/36 \text{ or } 1/18$ and it is the same for both.

Then, we are asked: "If Gus goes first what are his chances of being bought a drink?" This the chance that Gus wins - the probability that he rolls his favourite score. We don't know if his favourite score is 3 or 11, but we know the probability of rolling either 3 or 11 is equal to the sum of the probability of rolling  3 and the probability of rolling 11, that is[^2] 


$$P(3 \text{ or } 11) = P(3) + P(11) $$
and we know that $P(3) = P(11)= 1/18$. So the probability Gus wins is $2/18 = 1/9$. 

[^2]: This is usually called the addition rule. It states that for two events A and B, the probability that either one or both events occur is P(A or B) = P(A) + P(B) - P(A and B). The last term, P(A and B), is the probability that both events occur. In our case, it is zero because we cannot roll 3 and 11 at the same time, they cannot occur together when rolling a pair of dice. 


